Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-09-24 paint-contour-tree.tex
First pass done for upper bound

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

\begin{definition}
 A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
 one from the left and one from the right subtree of $v$.
\end{definition}

@@

\begin{definition}
\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
 a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
% 
% The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
\end{definition}

Fix $\chi(T)$, which fixes all the heaps $H_v$. We choose a special path decomposition, which can be related
to the running time of the algorithm. This is defined best as a subset of edges in $T$ such that each internal
vertex has degree exactly $1$. This naturally gives as path decomposition.
For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
 where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$.

@@

\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in P(T)} |p| \log |p|)$.
\end{theorem}

@@

Most of work goes into proving the following technical lemma.

@@

We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
which mostly imvolves routing calculations. Then we prove \Lem{cu-root}, where most
of the interesting work happens.

@@

\Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
%Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.

@@

\begin{lemma}
\label{lem:geometric}
 Let $p$ be the parent of $q$ in the shrub $\cU$. Then $|H_p|\geq \frac{3}{2}|H_q|$. 
\end{lemma}
\begin{proof} By construction, there is a heavy path (denoted $h(q)$) that is a child of $p$ in $\pmax(T)$,
and an ancestor of $q$. The chain of ancestors from $q$ to $h(q)$ consists only of heavy paths.
These paths contain at most $|H_p|/100$ vertices of $H_p$, so $|H_{h(q)}| \geq 99|H_p|/100$.

Consider the base
of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
Their parent is called $u$. Note that both $u$ and $v$ are nodes in the path $p$.
Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 1$,
$|H_u| \geq 2|H_w| - 1$. Let $b$ be the base of $p$.
We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a light path, $|p| < \sqrt{|H_p|}/100$.
Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for small constant $\delta > 0$).
Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| > 99|H_q|/100$,
to complete the proof.

%
%
% Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
% i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
% As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
% the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
% Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
% Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
% Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
\end{proof}

This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires much more work.
It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$.
Furthermore, as a leaf path in $p$, some leaf
$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is the
union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
for a child $q'$ in $\cU$.

@@

\begin{proof} (of \Lem{support})
\end{proof}

>>>>

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

\begin{definition}
 A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
 one from the left child and one from the right child subtree of $v$.
\end{definition}

@@

\begin{definition}
\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
 a set of vertex disjoint leaf paths.  
% 
% The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
\end{definition}

Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
Fix $\chi$, which fixes all the heaps $H_v$. We choose a special path decomposition that is best defined as a subset of edges in $T$ such that each internal
vertex has degree at most $2$. This naturally gives a path decomposition.
For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
 where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$.

@@

\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
\end{theorem}

@@

Most of the work goes into proving the following technical lemma.

@@

We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.

@@

\Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
We will work with the \emph{reduced} heap $\redH_p$. This is the subset of $H_p$
of vertices that only appear in light paths. By definition, for any unsupported light path
(hence, any non-root $p \in \cU$), $|H'_p| \geq 99|H_p|/100$. For notational convenience, we simply
set $H'_r = H_r$.
%Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.

@@

\begin{lemma}
\label{lem:geometric}
 Let $p$ be the parent of $q$ in the shrub $\cU$. Then $|\redH_p|\geq \frac{3}{2}|\redH_q|$. 
\end{lemma}
\begin{proof} By construction, there is a heavy path (denoted $h(q)$) that is a child of $p$ in $\pmax(T)$,
and an ancestor of $q$. The chain of ancestors from $q$ to $h(q)$ consists only of heavy paths.
Since $q$ is unsupported, these paths contain at most $|H_p|/100$ vertices of $H_p$. Thus, $|H_{h(q)}| \geq 99|H_p|/100$.

Consider the base
of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
Their parent is called $u$. Note that both $u$ and $v$ are nodes in the path $p$.
Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 1$,
$|H_u| \geq 2|H_w| - 1$. Let $b$ be the base of $p$.
We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a light path, $|p| < \sqrt{|H_p|}/100$.
Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for small constant $\delta > 0$).
Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| > 99|H_q|/100$
to get $|H_p| \geq 197|H_q|/100$. The proof is completed by noting that $|\redH_p| \geq 99|H_p|/100$ and $|H_q| \geq |\redH_q|$.
%
%
% Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
% i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
% As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
% the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
% Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
% Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
% Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
\end{proof}

This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires much more work.
It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$.
Furthermore, as a leaf path in $p$, some leaf
$\ell$ of $T$ in one of the ends of $p$. The reduced heap $\redH_p$ is the
union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
for a child $q'$ in $\cU$.

We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
By definition, $|\redH_p| = \sum_{q \in \cU_p} R_p$. 
Note that $R_p$ can be negative. 
Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_p$.
Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.

\begin{claim} \label{clm:loss} $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$
\end{claim}

\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} R^+_q - R_q$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
Traverse $\pmax(T)$ from $q'_i$ to $q$.
Other than $q$, all other nodes encountered are heavy paths and hence
do not affect reduced heaps.
The vertices of $\redH_{q'_i}$ that are deleted
are exactly those present in the path $q$ (which are paths in $T$) traversed. 
Any vertex in $q$ can be deleted from at most two reduced heaps $\redH_{q'_i}$.
When $R_q$ is negative, it is at most $2|q|$.
We sum over all $q$ to complete the proof.
\end{proof}

\ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
$D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.

 and traverse $\pmax(T)$ up to $p$. Suppose we encounter
the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
are removed from $H_p$ are exactly the vertices that lie in these paths. 
We split this contribution into heavy and light paths, observing that the latter
are all unsupported. Hence, we can (crudely) upper bound the number of vertices
removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
}

The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
We stress that the all the previous work is mostly the setup for this claim.

\begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
\end{claim}

\begin{proof} Since $q$ is an unsupported light path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
by a charge redistribution scheme. Assume that each $q \in \cU_p$
starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.

After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
$q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i$.
$$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
\end{proof}

We wrap it all up by proving part (ii) of \Lem{cu-root}.

\begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
\end{claim}

\begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.

We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
Suppose not. Let $p, p'$ be the counterexample furthest from the root,
where $p$ is the parent of $p'$. 
Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
That completes the whole proof.
\end{proof}

@@

\section{Ben's old stuff}
